Decompositions of congruence subgroups \\ of Chevalley groups

### 2018-02-13 paper.tex
Restored old statement of DV; swapped DV and Gauss; slicker proof of DV

\usepackage[hyperref=true, backend=bibtex, citestyle=numeric-comp, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}            
\addbibresource{paper.bib}
\renewbibmacro*{volume+number+eid}{\ifentrytype{article}{\- \iffieldundef{volume}{}{Vol.~\printfield{volume},}\iffieldundef{number}{}{ No.~\printfield{number},}}}
\renewbibmacro{in:}{\ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
\newbibmacro{string+doi}[1]{\iffieldundef{doi}{\iffieldundef{url}{#1}{\href{\thefield{url}}{#1}}}{\href{http://dx.doi.org/\thefield{doi}}{#1}}}
\DeclareFieldFormat[article, inproceedings, inbook, thesis]{title}{\usebibmacro{string+doi}{\mkbibquote{#1}}}
\renewcommand*{\bibfont}{\footnotesize}

@@

\section{Introduction}\label{sec:intro}
The aim of this paper is to generalize several decompositions of Chevalley groups over rings 
 to their principal congruence subgroups and relative elementary subgroups.
We focus on the so-called ``parabolic decompositions'', i.e. decompositions involving parabolic and unipotent factors. 
One of our main results is the following theorem.
\begin{thm}[Dennis--Vaserstein decomposition]\label{thm:DennisVaserstein}
Let $\Phi$ be an irreducible root system of rank $\ell\geqslant 2$ and $\{ r, s \}$ be a pair of distinct endnodes of the Dynkin diagram of $\Phi$.
Denote by $d$ the distance between $r$ and $s$ on the Dynkin diagram.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\asr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{1, \ell \}$,
 \item $\sr(I) \leqslant d$ otherwise.
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$
\end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
The condition $\asr(I) \leq d$ is a relative version of the \emph{absolute stable range} condition of D.~Estes and J.~Ohm, which depends only on the ideal $I$ (see section~\ref{sec:rel-asr}).
Another results of the paper are relative analogues of Gauss and Bass--Kolster decompositions (see~\cref{thm:srRI1,thm:BassKolster}).

@@

Before we proceed with the proof of~\cref{thm:DennisVaserstein} let us remark that in the special case $\ell=2$
 $\E(\Phi, R, I)$ admits Gauss decomposition (see~\cref{thm:srRI1} below).
It is not hard to see that Dennis--Vaserstein decomposition follows from Gauss decomposition, 
 therefore, in the sequel we may assume, without loss of generality, that $\ell > 2$.

@@

\begin{table}[htb]
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\quad} c @{\qquad} c @{\qquad} c @{\qquad} c}
\Phi                                 & s    &r      & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\Delta_r$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell, \ \ell\geqslant 2           & 1    &\ell   & \ell+1         & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rB_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell+1        & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rC_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\
\rD_\ell, \ \ell\geqslant 4           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 2  \\ 
\rD_\ell, \ \ell\geqslant 4           & \ell &\ell-1 & 2^{\ell-1}     & \text{half-spinor}   & \rA_{\ell-1}              & \ell           & \ell-2  \\
\rE_\ell, \ \ell=6, 7, 8              & \ell &2      & 27, 56, 248    & \text{minimal}       & \rA_{\ell-1}              & \ell           & 3       \\ 
\rE_\ell, \ \ell=6, 7, 8              & \ell &1      & 27, 56, 248    & \text{minimal}       & \rD_{\ell-1}              & 2(\ell-1)      & \ell-1 \\
\rE_\ell, \ \ell=6, 7, 8              & 1    &2      & 27, 133, 3875  & \text{???}           & \rA_{\ell-1}              & ?              & ? \\
\rF_4,                                & 4    & 1     & 26             & \text{minimal}       & \rC_{3}                   & 6              & 3 \end{array}\]
 \caption{List of representations used in the proof of \cref{thm:DennisVaserstein}.} \label{table:dv-reps}
\end{table}
The key step in the proof of~\cref{thm:DennisVaserstein} is to verify that the normalizer $N(A_{rs})$ is suffiently large.
This is accomplished in a series of lemmas. The following lemma identifies the ``obvious'' elements of $N(A_{rs})$.
\begin{lemma}\label{lemma:dv-normal} 
For every $\alpha \in \Delta_{r, s} \cup \Phi^+ $ one has $X_\alpha(R) \subseteq N(A_{rs})$. \end{lemma}
\begin{proof}
Notice that for every $i$ the group $\EP_i(R, I)$ is normalized by $\EP_i(R)$, hence it is normalized by $X_\alpha(R)$, $\alpha \in S_i^+$.
Since $\E(\Delta_{r, s}, R)$ normalizes $\U(\Sigma_r^- \cap \Sigma_s^-, I)$, we obtain the required assertion for $\alpha \in \Delta_{r, s}$.

@@

\begin{proof} [Proof of \cref{thm:DennisVaserstein}]
From Lemmas~\ref{lemma:Stein_reduction} and~\ref{lemma:dv-normal} it follows that $N(A_{rs})$ contains $\EP_s(R)$, 
 hence by~\cref{lem:LN} we obtain that $\EP_s(R, I) \subseteq L(A_{rs})$.

Our next goal is to verify the inclusion $\U(\Sigma_s^-, I) \subseteq L(A_{rs})$. Our main tools will be~\cref{lem:LN} and \cref{item-trans2}.
Since $\widetilde{W}(\Delta_s) \subseteq \E(\Delta_s, R) \subseteq N(A_{rs})$ and 
 $X_{\alpha}(I) \subseteq L(A_{rs})$ for $\alpha \in \Sigma^-_s\cap \Delta_r$, 
we immediately obtain that $L(A_{rs})$ contains root subgroups $X_\alpha(I)$ for all $\alpha$ lying in the orbit $O = W(\Delta_s) \cdot (\Sigma^-_s\cap \Delta_r)$.
Thus, we are left to consider five cases when $O$ does not coincide with $\Sigma^-_s$.
\begin{enumerate} 
 \item \textit{Case $\Phi = \rE_8$, $s=8$, $r=1,2$.}  
 There is only one root in $\Sigma_8^-$ not lying in $O$, namely $-\alpha_{\max}$ with $\shape(\{\alpha_8\}, -\alpha_{\max}) = -2$.
 Since $X_{-\alpha_8}(R) \subseteq N(A_{rs})$, $X_{\alpha_8-\alpha_{\max}}(I) \subseteq L(A_{rs})$ the desired inclusion
  follows from Chevalley commutator formula $x_{\alpha_{\max}}(\pm ab)=[x_{\alpha_8-\alpha_{\max}}(a), x_{\alpha_8}(b)].$
 \item \textit{Case $\Phi = \rE_7,$ (resp. $\Phi= \rE_8$), $s=1$, $r=2$.}
 There is only one (resp. there are $14$) roots in $\Sigma_1^-$ with $1$-shape $-2$, which are not contained in $O$ for this reason.
 Denote by $\beta$ the maximal root of $\Phi$ (resp. the maximal root of $\Delta_8 \cong \rE_7$).
 It is easy to check that $\shape(\{\alpha_1\}, \beta) = 2$ but $-\beta+\alpha_1 \in O$, therefore the assertion can be obtained via the same argument. 
 \item \textit{Case $\Phi = \rB_\ell$, $s=1$, $r=\ell$.} 
 Only the sole short root of $\Sigma_1^-$ is not contained in $O$. Consider the following two roots of $\Phi$:
 \begin{equation}\nonumber \beta = \alpha_2 + \ldots + \alpha_\ell,\ \gamma = -\alpha_1 - 2\alpha_2 - \ldots -2\alpha_\ell.\end{equation}
 We need to verify $X_{\beta+\gamma}(I) \subseteq L(A_{rs})$, but this follows from $X_{\gamma}(I) \cup X_{-\alpha_1}(I) \subseteq L_{rs}$,
  $X_{\beta}(R) \subseteq N(A_{rs})$ using Chevalley commutator formula:
  \begin{equation} \label{eq:CCF} x_{\beta+\gamma}(-ab) = [x_{\gamma}(a),\, x_{\beta}(b)] \cdot x_{\gamma + 2\beta}(ab^2). \end{equation}
 \item \textit{Case $\Phi = \rF_4$, $s=4$, $r=1$.} There are three possibilities for the $s$-shape and length of $\alpha\in \Sigma_4^-$: 
 \begin{itemize}
  \item $\shape(\{\alpha_4\}, \alpha) = 1$, $\alpha$ is short (there are 8 such roots);
  \item $\shape(\{\alpha_4\}, \alpha) = 2$, $\alpha$ is long (there are 6 such roots);
  \item $\shape(\{\alpha_4\}, \alpha) = 2$, $\alpha$ is short (there is only one such root: $\delta = -\alpha_1 -2\alpha_2-3\alpha_3-2\alpha_4$).
 \end{itemize}
 Since $\Delta_r \cap \Sigma_s^-$ contains roots of the first two types, we only need to verify $X_\delta(I) \subseteq L(A_{rs})$. 
 But the latter inclusion follows from formula~\eqref{eq:CCF} with $\beta = \alpha_2 + \alpha_3,\ \gamma = -\alpha_{{\max}}$. 
 \item \textit{Case $\Phi = \rC_\ell$, $s=1$, $r=\ell$.} 
  In this case the inclusion $X_{-\alpha_{\max}}(I) \subseteq L(A_{rs})$ does not follow from commutator formulae, therefore we need to invoke stability assumptions one more time.
  From~\cref{lemma:DVST} below we obtain that $N(A_{rs})$ contains $X_{-\alpha_1}(R)$ and therefore coincides with $\E(\Phi, R)$.
  The required inclusion now follows~\cref{item-trans1}.
\end{enumerate}

Thus, we have just shown that $\mathcal{X}\subseteq L(A_{rs})$. 
Recall from~\cite[Theorem~3.4]{S} that the group $\E(\Phi, R, I)$ is generated by the subset $\mathcal{Z}(\Sigma_s^-)$.
On the other hand, from $\mathcal{Z}(\Sigma_s^-) \subseteq \mathcal{X}^{\EP_s(R)} \subseteq L(A_{rs})^{N(A_{rs})} \subseteq L(A_{rs})$ we conclude that 
 $\E(\Phi, R, I) \subseteq L(A_{rs})$, which completes the proof.
\end{proof}

\begin{lemma}\label{lemma:DVST}
Assume that $\Phi=\rC_\ell$ and $\sr(I) \leq \ell-1$, then $X_{-\alpha_s}(R) \subseteq N_{rs}$.
\end{lemma}
\begin{proof}
Denoteb by $C$ the set consisting of elements $g \in \EP_s(R, I)$ for which matrix entries $(g_{i, 2})$, $i=2, \ldots, \ell$ form an $I$-unimodular column of height $\ell-1$

>>>>

\usepackage[hyperref=true, backend=bibtex, citestyle=numeric-comp, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}            
\addbibresource{paper.bib}
\renewbibmacro*{volume+number+eid}{\ifentrytype{article}{\- \iffieldundef{volume}{}{Vol.~\printfield{volume},}\iffieldundef{number}{}{ No.~\printfield{number},}}}
\renewbibmacro{in:}{\ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
\newbibmacro{string+doi}[1]{\iffieldundef{doi}{\iffieldundef{url}{#1}{\href{\thefield{url}}{#1}}}{\href{https://doi.org/\thefield{doi}}{#1}}}
\DeclareFieldFormat[article, inproceedings, inbook, thesis]{title}{\usebibmacro{string+doi}{\mkbibquote{#1}}}
\renewcommand*{\bibfont}{\footnotesize}

@@

\section{Introduction}\label{sec:intro}
The aim of this paper is to generalize several decompositions of Chevalley groups over rings 
 to their principal congruence subgroups and relative elementary subgroups.
We focus on the so-called ``parabolic decompositions'', i.e. decompositions involving parabolic and unipotent factors. 
One of our main results is the following theorem.
\begin{thm}[Dennis--Vaserstein decomposition]\label{thm:DennisVaserstein}
Let $\Phi$ be an irreducible root system of rank $\ell\geqslant 2$ and $\{ r, s \}$ be a pair of distinct endnodes of the Dynkin diagram of $\Phi$.
Denote by $d$ the distance between $r$ and $s$ on the Dynkin diagram.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\sr(I) \leqslant d$ for $\Phi$ classical or $\Phi=\rG_2, \rF_4$;
 \item $\sr(I) \leqslant d$ for $\Phi=\rE_6,\rE_7$ with $\{r, s\} = \{2, \ell\}$;
 \item $\asr(I) \leqslant d$ for $\Phi=\rE_6,\rE_7$ with $\{r, s\} = \{1, \ell \}$, 
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$
\end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
The condition $\asr(I) \leq d$ is a relative version of the \emph{absolute stable range} condition of D.~Estes and J.~Ohm, which depends only on the ideal $I$ (see section~\ref{sec:rel-asr}).
Another results of the paper are relative analogues of Gauss and Bass--Kolster decompositions (see~\cref{thm:srRI1,thm:BassKolster}).

@@

Before we proceed with the proof of~\cref{thm:DennisVaserstein} let us remark that in the special case $\ell=2$, $\sr(I)\leq 1$
 we already known that $\E(\Phi, R, I)$ admits Gauss decomposition.
It is not hard to see that Dennis--Vaserstein decomposition follows from Gauss decomposition, 
 therefore, in the sequel we may assume, without loss of generality, that $\ell > 2$.

@@

\begin{table}[htb]
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\quad} c @{\qquad} c @{\qquad} c @{\qquad} c}
\Phi                                  & s    &r      & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\Delta_r$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell, \ \ell\geqslant 2           & 1    &\ell   & \ell+1         & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rB_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell+1        & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rC_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\
\rD_\ell, \ \ell\geqslant 4           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 2  \\ 
\rD_\ell, \ \ell\geqslant 4           & \ell &\ell-1 & 2^{\ell-1}     & \text{half-spinor}   & \rA_{\ell-1}              & \ell           & \ell-2  \\
\rE_\ell, \ \ell=6, 7                 & \ell &2      & 27, 56         & \text{minimal}       & \rA_{\ell-1}              & \ell           & 3       \\ 
\rE_\ell, \ \ell=6, 7                 & \ell &1      & 27, 56         & \text{minimal}       & \rD_{\ell-1}              & 2(\ell-1)      & \ell-1 \\
\rF_4,                                & 4    & 1     & 26             & \text{minimal}       & \rC_{3}                   & 6              & 3 \end{array}\]
 \caption{List of representations used in the proof of \cref{thm:DennisVaserstein}.} \label{table:dv-reps}
\end{table}
The key step in the proof of~\cref{thm:DennisVaserstein} is to verify that the normalizer $N(A_{rs})$ is suffiently large.
This is accomplished in a series of lemmas. The following lemma identifies the ``obvious'' elements of $N(A_{rs})$.
\begin{lemma}\label{lemma:dv-normal} 
For every $\alpha \in \Delta_{r, s} \cup \Phi^+ $ one has $X_\alpha(R) \subseteq N(A_{rs})$. \end{lemma}
\begin{proof}
Notice that for every $i$ the group $\EP_i(R, I)$ is normalized by $\EP_i(R)$, hence it is normalized by $X_\alpha(R)$, $\alpha \in S_i^+$.
Since $\E(\Delta_{r, s}, R)$ normalizes $\U(\Sigma_r^- \cap \Sigma_s^-, I)$, we obtain the required assertion for $\alpha \in \Delta_{r, s}$.

@@

We will need one more root-theoretic lemma before we proceed with the proof of~\cref{thm:DennisVaserstein}.
Let $\Phi$, $r$ and $s$ be as in~\cref{table:dv-reps}.
Denote by $O$ the subset of roots of $\Sigma_s^-$ lying in the orbit $W(\Delta_s) \cdot (\Sigma^-_s\cap \Delta_r)$.
\begin{lemma} \label{lemma:root-lemma}
Suppose that $\Phi \neq \rC_\ell$. The following alternative holds: either $O$ coincides with $\Sigma_s^-$ or
$\Phi=\rB_\ell, \rF_4$ and $\Sigma_s^- \setminus O$ consists of the sole short root $\gamma$, which can be presented as a sum 
 $\gamma = \alpha + \beta$ for some short root $\alpha \in \Phi^+$ and some long root $\beta \in O$ satisfying $\beta + 2\alpha \in O$. 
\end{lemma}
\begin{proof}
From \cref{lemma:abs} it follows that $O$ does not coincide with $\Sigma^-_s$ only in the following two cases.
 \begin{enumerate} 
 \item \textit{Case $\Phi = \rB_\ell$, $s=1$, $r=\ell$.} 
 Only the sole short root of $\Sigma_1^-$ is not contained in $O$, denote it by $\gamma$. Set $\alpha = \alpha_2 + \ldots + \alpha_\ell,\ \beta = -\alpha_1 - 2\alpha_2 - \ldots -2\alpha_\ell.$
 It is clear that $\alpha \in \Phi^+$, $\beta, \beta + 2\alpha \in O$.
 \item \textit{Case $\Phi = \rF_4$, $s=4$, $r=1$.} There are three possibilities for the $s$-shape and length of $\alpha\in \Sigma_4^-$: 
 \begin{itemize}
  \item $\shape(\{\alpha_4\}, \alpha) = 1$, $\alpha$ is short (there are 8 such roots);
  \item $\shape(\{\alpha_4\}, \alpha) = 2$, $\alpha$ is long (there are 6 such roots);
  \item $\shape(\{\alpha_4\}, \alpha) = 2$, $\alpha$ is short (there is only one such root, denote it $\gamma$).
 \end{itemize}
 Since $\Delta_r \cap \Sigma_s^-$ contains roots of the first two types, only $\gamma$ is not contained in $O$.
 Clearly, $\gamma = -\alpha_1 -2\alpha_2-3\alpha_3-2\alpha_4 = \alpha + \beta$ for $\alpha = \alpha_2 + \alpha_3\in \Phi^+,$ $\beta = -\alpha_{{\max}} \in O$. \qedhere
\end{enumerate}
\end{proof}

\begin{proof} [Proof of \cref{thm:DennisVaserstein}]
From Lemmas~\ref{lemma:Stein_reduction} and~\ref{lemma:dv-normal} it follows that $N(A_{rs})$ contains $\EP_s(R)$.
Thus, by~\cref{lem:LN} we obtain that $\EP_s(R, I) \subseteq L(A_{rs})$.

Our next goal is to verify the inclusion $\U(\Sigma_s^-, I) \subseteq L(A_{rs})$.
Since $\widetilde{W}(\Delta_s) \subseteq \E(\Delta_s, R) \subseteq N(A_{rs})$ and 
 $X_{\alpha}(I) \subseteq L(A_{rs})$ for $\alpha \in \Sigma^-_s\cap \Delta_r$, 
we obtain from~\cref{lem:LN} and \cref{item-trans2} that $L(A_{rs})$ contains root subgroups $X_\alpha(I)$ for $\alpha \in O = W(\Delta_s) \cdot (\Sigma^-_s\cap \Delta_r)$.
Thus, by~\cref{lemma:root-lemma} we are left to consider two cases when $O$ does not coincide with $\Sigma^-_s$. 
\begin{enumerate} 
  \item \textit{Case $\Phi = \rB_\ell, \rF_4$.} Let $\alpha, \beta, \gamma$ be as in~\cref{lemma:root-lemma}.
  Since $X_{\beta}(I), X_{\beta+2\alpha}(I) \subseteq L(A_{rs})$ and $X_{\alpha}(R) \subseteq N(A_{rs})$ we obtain the required inclusion
   $X_{\gamma}(I)$ using Chevalley commutator formula:
   \begin{equation} \nonumber x_{\gamma}(\pm ab) = [x_{\beta}(a),\, x_{\alpha}(b)] \cdot x_{\beta + 2\alpha}(ab^2). \end{equation}
 \item \textit{Case $\Phi = \rC_\ell$, $s=1$, $r=\ell$.} 
  In this case we need to invoke stability assumptions one more time.
  From~\cref{lemma:DVST} below we obtain that $N(A_{rs})$ contains $X_{-\alpha_1}(R)$ and therefore coincides with $\E(\Phi, R)$.
  The required inclusion now follows~\cref{item-trans1}.
\end{enumerate}
Thus, we have shown that $\mathcal{X}\subseteq L(A_{rs})$. 
Recall from~\cite[Theorem~3.4]{S} that the group $\E(\Phi, R, I)$ is generated by the subset $\mathcal{Z}(\Sigma_s^-)$.
On the other hand, from $\mathcal{Z}(\Sigma_s^-) \subseteq \mathcal{X}^{\EP_s(R)} \subseteq L(A_{rs})^{N(A_{rs})} \subseteq L(A_{rs})$ we conclude that 
 $\E(\Phi, R, I) \subseteq L(A_{rs})$, which completes the proof.
\end{proof}

\begin{lemma}\label{lemma:DVST}
Assume that $\Phi=\rC_\ell$ and $\sr(I) \leq \ell-1$, then $X_{-\alpha_s}(R) \subseteq N_{rs}$.
\end{lemma}
\begin{proof}
Denote by $C$ the set consisting of elements $g \in \EP_s(R, I)$ for which matrix entries $(g_{i, 2})$, $i=2, \ldots, \ell$ form an $I$-unimodular column of height $\ell-1$
